---
title: "Reproducible Research: Peer Assessment 1"
author: "Liz Anderson"
output: 
  html_document:
    keep_md: true
---
## Loading and preprocessing the data

``` {r library, echo=FALSE, results='hide', include=FALSE}
library(ggplot2)
library(dplyr)
options(scipen=999)
knitr::opts_chunk$set(fig.path="figure/")
```
``` {r files}
# download straight from the source
file <- file.path("activity.zip")
if (!file.exists(file)){
  url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
  download.file(url, file, method = "curl")
}

# unzip file into data folder
if (file.exists(file)){
  unzip(file, exdir='./data')
}

# prep data, need date column
data_na <- read.csv(file.path("./data","activity.csv"))
data <- na.omit(data_na)
data$date <- as.Date(data$date)
```


## What is mean total number of steps taken per day?
``` {r steps_per_day}
steps_per_day <- aggregate(steps ~ date, data, sum)
ggplot(steps_per_day, aes(x=steps)) + geom_histogram(position = "identity", binwidth = 300) +
  scale_x_continuous(breaks = round(seq(0, 22000, by = 2500),1)) +
  ggtitle("Steps per day") + 
  labs(x = "Steps per day", y = "Count")

mean_steps <- ceiling(mean(steps_per_day$steps))
median_steps <- median(steps_per_day$steps)
```
The mean steps per day is `r mean_steps`.

The median steps per day is `r median_steps`.


## What is the average daily activity pattern?
``` {r avg_daily}
# Merge with your date
data$time <- as.POSIXct(paste(data$date, " ", sprintf("%04s", data$interval)), format = "%Y-%m-%d %H%M")
data$num_interval <- as.numeric(data$interval)

# aggreate the intervals of the day by function
mean_step_by_int <- aggregate(steps ~ num_interval, data, mean)
median_step_by_int <- aggregate(steps ~ num_interval, data, median)

# interval of max number of steps
max_mean_int <- mean_step_by_int[max(mean_step_by_int$steps),]$num_interval

ggplot(mean_step_by_int, aes(x = num_interval, y = steps)) + geom_line() +
  scale_x_continuous(breaks = round(seq(0, 2355, by = 200),1)) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  xlab(label = "Time (24hr)")
```

On average, `r max_mean_int` was the interval of time (24hr) with the most steps to occur during the day.


## Imputing missing values
``` {r impute_hide_1, echo=FALSE, results='hide', include=FALSE}
data_impute <- data_na
nas_before <- sum(is.na(data_impute$steps))
```

```{r impute_plot}
# impute NAs in by replacing with the mean of the interval of all the days
for (i in 1:nrow(data_impute)){
  data_impute[i,"steps"] <- ifelse(is.na(data_impute[i,"steps"]), 
    mean_step_by_int$steps[mean_step_by_int$interval==data_impute$interval[i]],data_impute[i,"steps"])
}

# agg data to plot
data_impute_agg <- aggregate(steps ~ interval, data_impute, sum)
ggplot(data_impute_agg, aes(x=steps)) + geom_histogram(position = "identity", binwidth = 300) +
  scale_x_continuous(breaks = round(seq(0, 22000, by = 2500),1)) +
  ggtitle("Imputed Data - replace with mean") + 
  labs(x = "Steps", y = "Count")
```

``` {r impute_hide_2, echo=FALSE, results='hide', include=FALSE}
mean_steps_impute <- ceiling(mean(data_impute_agg$steps))
median_steps_impute <- median(data_impute_agg$steps)
```
The data was imputed by replacing the `r nas_before` NAs from the original dataset with the mean steps taken from the same time interval from the rest of the data.

The mean steps with <b>imputed data</b> per day is `r mean_steps_impute` compared to the mean of data removing NA values, `r mean_steps`, thats a difference of `r abs(mean_steps_impute-mean_steps)` steps.

The median steps  with <b>imputed data</b> per day is `r median_steps_impute` compared to the mean of data removing NA values, `r median_steps`, thats a difference of `r abs(median_steps_impute-median_steps)` steps.


## Are there differences in activity patterns between weekdays and weekends?
``` {r weekdays}
# create new column with weekdays
data$weekdays <- weekdays(data$date)

# create new column day that sets type of day
data <- mutate(data, day = ifelse(weekdays %in% c("Saturday", "Sunday"), "Weekend", "Weekday"))

# agg data to plot
days_agg <- aggregate(steps ~ day+num_interval, data, sum)
ggplot(days_agg, aes(x = num_interval, y = steps, color = day)) + geom_line() +
  facet_wrap(~day, nrow=2) +
  scale_x_continuous(breaks = round(seq(0, 2355, by = 200),1)) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  xlab(label = "Time (24hr)")
```
